Manuscript progress report

### 2024-01-16 Docs/NhanesApplExample.Rmd
time-varying variation of each PC in data application

set.seed(121)

@@

train_id <- unique(df$id)[1:Ntr]
train_df <- df %>% filter(id %in% train_id)

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
# keep 4 PCs
t2 <- Sys.time()
# t2-t1 
```

```{r PC}
data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
  rename(PC1 = X1, PC2=X2, PC3=X3,PX4=X4) %>%
  pivot_longer(2:5, names_to = "PC") %>%
  ggplot()+
  geom_line(aes(x=t, y=value, col = PC))+
  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
```

@@

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1
```

I ran into a few problems at this step. Not only does the mgcv::bam model take a really long time to fit, the re-evaluated eigenvalues are not proportional to the FPCA eigenvalues. They also lost the decreasing nature. Specifically:

@@

The FPCA eigenvalues, as expected, decreases from PC1 to PC4. However after re-evaluation, the eigenvalue corresponding to PC2 is less than PC3. Also, the proportion of these two set of estimates is not constant, as below:

@@

>>>>

set.seed(1114)

@@

train_id <- unique(df$id)[1:Ntr] # training sample size = 1000
train_df <- df %>% filter(id %in% train_id)

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
# keep 3 PCs
t2 <- Sys.time()
# t2-t1

# fpca_mod$evalues/sum(fpca_mod$evalues)
```

```{r PC}
data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1 = X1, PC2=X2, PC3=X3, PC4=X4) %>%
  pivot_longer(2:5, names_to = "PC") %>%
  ggplot()+
  geom_line(aes(x=t, y=value, col = PC))+
  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
```

@@

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1 # 7.5 minutes
```

It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically

@@

Also, the proportion of these two set of estimates is not constant, as below:

@@

### Percentage of variance explained by each PC

We now that the variance of individual-specific random effects, in theory, is:

\[\begin{aligned}
Var(b_i(t)) & = Var(\sum_{k=1}^K \xi_{ik}\phi_k(t)) \\
& = \sum_{k=1}
^K\phi_k(t)^2\lambda_k
\end{aligned}\]

Here I'd like to compare the variantion for each PC calculated using estimated from the **biased** FPCA estimates and the supposedly **debiased** from the dibased GLMM model.

```{r var_bi_fpca}
df_pc_fpca <- data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>% 
  mutate(eval1=fpca_mod$evalues[1], 
         eval2=fpca_mod$evalues[2], 
         eval3=fpca_mod$evalues[3], 
         eval4=fpca_mod$evalues[4]) %>%
  mutate(var1 = eval1*PC1^2, var2 = eval2*PC2^2, var3 = eval3*PC3^2, var4 = eval4*PC4^2)

```{r var_bi_glmm}
# re-evaluated eigenvalues
evals_new <- 1/debias_glmm$sp[2:5]

df_pc_glmm <- df_phi %>%
  mutate(eval1=evals_new[1], 
         eval2=evals_new[2], 
         eval3=evals_new[3], 
         eval4=evals_new[4]) %>%
  mutate(var1 = eval1*phi1^2, var2 = eval2*phi2^2, var3 = eval3*phi3^2, var4 = eval4*phi4^2)
```

```{r pc_var}
df_pc_fpca$est <- "FPCA"
df_pc_glmm$est <- "Debias"

bind_rows(
  df_pc_fpca %>% select(t, starts_with("var"), est),
  df_pc_glmm %>% select(sind, starts_with("var"), est) %>% rename(t=sind)
) %>% 
  rename(PC1=var1, PC2=var2, PC3=var3, PC4=var4) %>%
  pivot_longer(2:5, names_to = "PC", values_to = "variance") %>%
  ggplot()+
  geom_line(aes(x=t, y=variance, col=est, group=est))+
  facet_wrap(~PC)+
  labs(title = "Time-varying variation of each PC")

Things I see from the plots above:

1. PC1 basically has no effect in the earlier track. It explains variation later during the day
2. PC2 seems to work primary at night and a bit earlier time interval from PC1
3. PC2 and PC3 seems to have similar shape? In fact, I think PC4 has a similar shape as well

I am also curious about how the individual scores change after debias

```{r score_df}
df_score_fpca <- data.frame(id = 1:Ntr, fpca_mod$scores) %>%
  pivot_longer(2:5, names_to = "PC", values_to = "score") %>%
  mutate(est = "FPCA")

df_score_glmm <- data.frame(
  id = 1:Ntr,
  X1 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi1.")) %>% t(),
  X2 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi2.")) %>% t(),
  X3 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi3.")) %>% t(),
  X4 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi4.")) %>% t()
  ) %>%
  pivot_longer(2:5, names_to = "PC", values_to = "score") %>%
  mutate(est = "Debias")

```{r}
bind_rows(df_score_fpca, df_score_glmm) %>%
  ggplot(aes(x=est, y=score, col=est))+
  geom_boxplot()+
  geom_jitter(size = 0.2, alpha=0.5)+
  facet_wrap(~PC, scales = "free_y")+
  labs(title = "Score of each PC")

### 2024-01-16 Docs/manuscript.Rmd
time-varying variation of each PC in data application

We would like to use a second reference method for predictive performance comparision. This would be a generalized function-on-function regression where the history records are used to predict a equal length of outcome. Formulation as follows:

$$g(E[Y_i(t)]) = \beta_0(t) + \beta_1 (t)Y_i(s),\ s<t$$

In this case, the outcome must be evaluated at the same time points as the covariate. That is, s and t must have the same number of measurement points. For example, if t is (0, 0.2], then s can be either (0.2, 0.4), (0.4, 0.6), (0.6, 0.8) or (0.8, 1).

What to do when the history t is (0, 0.4)? How share we prediction on 0.2 windows including (0.4, 0.6), (0.6, 0.8) or (0.8, 1)? Does this mean we do not utilize the full observed track but instead only take the most recent subtrack (0.2, 0.4)? Such scenarios remains unclear.

As for the model fitting, I am not sure how to construct a function-on-function regression using bam or gam. While most functions I found in R does not account for the "generalized" part. At first glance, the model framework is the same as a regular functional regression model:

$$g(E[Y_i(t)]) = \sum_{k}\xi_{k}\phi_k(t) + Y_i(t)\sum_{l}\zeta_{l}\phi_l(s),\ s<t$$

so I fit a gam model where history is [0, 0.2) and future is [0.2, 0.4).

```{r data_format, message=FALSE}
load(here("Data/sim_data.RData"))

# Let's start with a one-simulation example
df_m <- sim_data[[1]]
# class(df_m$Y)
# df_m$Y <- as.factor(df_m$Y)

# observation/prediction window 
windows <- seq(0, 1, by = 0.2)
df_m$window <- cut(df_m$t, breaks=windows, labels = 1:5,
                       include.lowest = T)

# split data
N_train <- 500
N_test <- 100
train_df <- df_m %>% filter(id %in% 1:N_train)
test_df <- df_m %>% filter(!id %in% 1:N_train)

# put covariates and outcome together
dfm1_tr <- train_df %>% filter(window==2) %>% 
  rename(Yf=Y, tf=t) %>%
  left_join(train_df %>% filter(window==1) %>% select(id, sind, t, Y) %>% 
              mutate(sind=sind+200) %>%
              rename(Yh=Y, th = t))

dfm1_te <- test_df %>% filter(window==2) %>% 
  rename(Yf=Y) %>%
  left_join(test_df %>% filter(window==1) %>% select(id, sind, Y) %>% 
              mutate(sind=sind+200) %>%
              rename(Yh=Y))

# class(dfm1_tr$Yh)
```

```{r class.source='fold-show'}
fitm1 <- bam(Yf ~ s(tf, bs="cc", k=20) + s(tf, bs="cc", k=20, by = Yh) , 
                   family = binomial, data=dfm1_tr, 
                   method = "fREML",
                   discrete = TRUE)
summary(fitm1)
```

However, the prediction result of this model is very wired. Below is the in-sample estimation of latent process. I am not sure if this is because of a wrong design matrix or ignoring the individual random effect.

```{r}
dfm1_tr$eta_est <- predict(fitm1, newdata = dfm1_tr, type = "link")
dfm1_tr %>% filter(id %in% 1:4) %>% 
  mutate_at(vars(eta_i, eta_est), inv.logit) %>%
  ggplot()+
  geom_line(aes(x=tf, y=eta_i, col = "true"))+
  geom_line(aes(x=tf, y=eta_est, col = "pred"))+
  facet_wrap(~id)
```

Thinking about the two problems above, perhaps I would want to use the following framework instead. This framework doesn't require same density of s and t. But I am not sure how to construct the design matrix for this.

$$g(E[Y_i(t)]) = \beta_0(t) + \int_{s}\beta_1(t, s)Y_i(s), \ s<t$$

or

$$g(E[Y_i(t)]) = \beta_0(t) + F(Y_i(s), s, t), \ s<t$$

Perhaps I can try using [refund::ff](https://rdrr.io/cran/refund/man/ff.html)

```{r class.source='fold-show', error=TRUE}
library(refund)

th <- unique(dfm1_tr$th)
tf <- unique(dfm1_tr$tf)

Yhmat <- matrix(dfm1_tr$Yh, nrow=N_train, ncol = length(th), byrow = T)

Yhmat2 <- Yhmat[rep(1:N_train, each = length(th)), ]
Yf <- dfm1_tr$Yf
# length(Yf)
# dim(Yhmat2)

fitm1 <- fgam(Yf ~ s(tf, bs="cc", k=10) + af(Yhmat2, argvals=th) , 
                   family = binomial())
# I couldn't get this function to work
# I think for each observation in the future, all of the history is used as predictor
# therefore the whole historical function is repeatedly used for each subject?

>>>>

```{r}
load(here("Data/sim_data.RData"))
df_m <- sim_data[[1]]
df_m %>% filter(t>0.2 & t<=0.4)
```

We would like to use a second reference method for predictive performance comparision. In this madel, predictions on a specific interval are all made using the last few observations in the observed window. For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2($t_j: j=200, ... 200-(L-1)$) as time-fixed covariate to predict any future time.

Let's write out the model expression (with questionable notation). If the observations if up to $t_m$, then:

\[\begin{aligned}
g(E[Y_i(t)]) &= \beta_0(t) +\sum_{l=1}^L \beta_l (t)Y_i(t_l)\\
l & = m,...m-(j-1)\\
t &> t_m 
\end{aligned}\]

This is a simple function-on-scalar model with no random effect, meaning all subject with the same last observed outcome would have the same estimated/predicted latent track. This is clearly anti-intuitive. But adding random effects would make out-of-sample prediction impossible.

Under this framework, for each dataset we need to fit four models:

- Given 0-0.2, predict 0.2-1
- Given 0-0.4, predict 0.4-1
- Given 0-0.6, predict 0.6-1
- Given 0-0.8, predict 0.8-1

Below are some single dataset examples.

## L = 1

First let try to make prediction on an interval with only one historical observation (L = 1):

\[\begin{aligned}
g(E[Y_i(t)]) &= \beta_0(t) +\beta_1(t) Y_i(t_{m})\\
t &> t_m 
\end{aligned}\]

```{r data_format}
L <- 1 # number of observations to use as predictor

# prediction window 
windows <- seq(0, 1, by = 0.2)
df_m$window <- cut(df_m$t, breaks=windows, labels = 1:5,
                       include.lowest = T)

# split data
N_train <- 500
N_test <- 100
train_df <- df_m %>% filter(id %in% 1:N_train)
test_df <- df_m %>% filter(!id %in% 1:N_train)
```

```{r fit_model, class.source='fold-show'}
# format training data
## given [0, 0.2], predictor would be
y_obs_max <- train_df %>% filter(window==1) %>% arrange(desc(t)) %>%
  group_by(id) %>%
  slice(L) %>% select(id, Y) %>% rename(yl = Y)
## to predict (0.2, 0.4), outcome:
df_pred_tr <- train_df %>% filter(window!=1) %>%
  left_join(y_obs_max, by = "id") %>% 
  mutate_at(vars(Y, yl), as.factor)

# model
t1 <- Sys.time()
fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
                        s(t, bs="cc", k=20, by = yl),
                   family = binomial, data=df_pred_tr, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t2-t1

summary(fit_gen_fosr)
```

```{r est_train}
# estimation on the training set
df_pred_tr$eta_est <- predict(fit_gen_fosr, type = "link")

df_pred_tr %>%
  mutate_at(vars(eta_est, eta_i), function(x){exp(x)/(1+exp(x))}) %>%
  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
  filter(id %in% 1:4) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.5)+
  geom_line(aes(x=t, y=eta_est, col = "estimate"))+
  geom_line(aes(x=t, y=eta_i, col = "ture"))+
  facet_wrap(~id)+
  labs(title = "In-sample estimation")
```

```{r pred_test}
# format testing data
df_pred_te <- test_df %>% filter(window!=1) %>%
  left_join(test_df %>% 
              filter(window==1) %>% arrange(desc(t)) %>% group_by(id) %>%
              slice(L) %>% select(id, Y) %>% rename(yl = Y),
            by = "id") %>% 
  mutate_at(vars(Y, yl), as.factor)

# predict using the FOSR model fit above
df_pred_te$eta_est <- predict(fit_gen_fosr, newdata = df_pred_te, type = "link")

df_pred_te %>%
  mutate_at(vars(eta_est, eta_i), function(x){exp(x)/(1+exp(x))}) %>%
  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.5)+
  geom_line(aes(x=t, y=eta_est, col = "estimate"))+
  geom_line(aes(x=t, y=eta_i, col = "ture"))+
  facet_wrap(~id)+
  labs(title = "Out-of-sample prediction")
